A lottery model of density-dependent selection

### 2017-10-27 manuscript.tex
latest

Verbal interpretations of the $r$/$K$ scheme, and the logistic model, tend to obscure the generality of MacArthur's argument, and so it is worth briefly reviewing that argument here. Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$. By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not. This region locally corresponds to greater $n_1$ and/or $n_2$ than can be tolerated by type $1$, because $f_1$ and $f_2$ are strictly decreasing. In this sense, selection at high density favors the ability to keep growing at ever greater densities.

MacArthur then analyzed the possible outcomes of Eq.~\eqref{eq:macgeneral} in terms of "carrying capacities" (generally labeled ``$K$''), defined as the points where the nullclines intersect the axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$). These intersection points neatly summarize the high density behavior of each type provided that the nullclines are approximately linear. He concluded that in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}. Note that only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types. For instance, if $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, then  $K_{12}=\alpha_{12}$ is just the standard Lotka-Volterra competition coefficient. This led to the distinction between selection for greater saturation density ($K$-selection in the strict sense) and selection for greater competitive ability (``$\alpha$-selection'') at high density \citep{gill_1974,case_1974,joshi_2001}.

Yet the $r$/$K$ scheme is strikingly incompatible with the  model of selection

>>>>

Verbal interpretations of the $r$/$K$ scheme, and the widespread use of the logistic model when discussing it, have tended to obscure the compelling generality of MacArthur's argument. We therefore  briefly review that argument now. Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$. By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not. Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth of type $2$ in a region of greater $n_1$ and/or $n_2$ than can be tolerated by type $1$. The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$). These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, leading to the conclusion that in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}

Confusingly, only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first. Thus, we can distinguish between selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}. Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.

Yet  ``$K$-selection'' is incompatible with the most widely used model of selection.

there is a deeper issue with the $r$/$K$ scheme as a description of selection in crowded versus uncrowded conditions ---
